---
title: 'Lab 3 In-class assignment: Data summaries'
author: "brouwern@gmail.com"
date: "September 11, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Fisher's Iris dataset

Load it
```{r}
data(iris)
```

<br>

## What "is" iris?

There are several things listed, but "data.frame" is the important one
```{r}
is(iris)
```

<br>

### How many rows?
```{r}
#with dim: 1st number is rows
dim(iris)

#with nrow
nrow(iris)

```

<br>

## How many columns
```{r}
#with dim: 2nd number is columns
dim(iris)

#with ncol
ncol(iris)
```

<br>


## How many rows are numeric?  How many are categorical/grouping?
```{r}
summary(iris)
```

<br>


## What "is" the species column?
```{r}
is(iris$Species)
```

<br>

## How many "levels" does this factor have?
```{r}
levels(iris$Species)
```

<br>

## Mammal sleep data
```{r}
library(ggplot2)
data(msleep)
```

<br>

### What "is" msleep?

Several things listed, but "data.frame" is the most important one.
```{r}
is(msleep)
```

<br>

### How many rows?
```{r}
#with dim: 1st number is rows
dim(msleep)

#with ncrow
nrow(msleep)

```

<br>

### How many columns
```{r}
#with dim: 2nd number is columns
dim(msleep)

#with ncol()
ncol(msleep)
```

<br>


### How many rows are numeric?  How many are categorical/grouping?

For some reason the categorical variabls are in "character" (aka text) form.  (Note: I'm just showing a few of the columns)
```{r, results="hide"}
summary(msleep)
```

```{r, echo=FALSE}
summary(msleep)[,1:7]
```

### Make them explicitly categorical/factor
```{r}
msleep$genus <- factor(msleep$genus)
msleep$vore <- factor(msleep$vore)
msleep$order <- factor(msleep$order)
msleep$conservation <- factor(msleep$conservation)
```

Look at summary again
```{r, results="hide"}
summary(msleep)
```

```{r, echo=FALSE}
summary(msleep)[,1:4]
```


### Levels to genus variable

Output not shown; there are 77!
```{r, results="hide"}
levels(msleep$genus)
```


