---
title: "Doing Basic Math for Stats in R"
author: "Nathan L. Brouwer | brouwern@gmail.com"
date: "September 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 4, 
                      fig.height = 2,
                      warning = FALSE,
                      message = FALSE)

```

## Doing Math in R

* R can do everything a scientific calculator or spreadsheet can do.  * This includes basic functions like +, - , /, sqrt, etc, 
* Also basic functions like mean(), median(), sd() for the standard deviation and var() for the variance.
* In general in this course we will focus on having R do most of the calculations for our stats.  
* However, it's important to understand some of the underlying math.
* Today, to practice doing math in R and to learn about basic statsitcal functions, we'll do calculations  of the mean, variance, "sum of squares", and standard deviation "by hand" in R.  
* We'll then compare them to the output R produces.


## The USGS Breeding Bird Survey (BBS) data 

Below are the number of species observed during a recent year of surveys on all routes in PA

```{r, echo = F}
#route number w/in PA
route<- c(2,	3,	4,	5,	6,	8,	9,	10,	11,	12,	13,	14,	15,	16,	17,	18,	19,	20,	21,	23,	24,	25,	28,	29,	30,	31,	33,	34,	35,	36,	37,	39,	41,	42,	43,	46,	47,	48,	49,	50,	52,	53,	54,	56,	57,	58,	59,	61,	62,	63,	64,	67,	68,	69,	70,	71,	72,	76,	79,	83,	84,	85,	86,	88,	90,	91,	97,	107,	126,	127,	132,	151,	155,	165,	166,	173,	174,	177,	178,	180,	181,	182,	187,	189,	192,	193,	194,	195,	196,	275,	901,	902,	904,	905,	906,	907,	909,	910,	911)

#Number of species on each route
spp.rich	<-c(69,	58,	69,	65,	65,	61,	74,	57,	61,	61,	56,	40,	67,	56,	58,	69,	73,	59,	63,	51,	77,	71,	59,	77,	63,	64,	74,	74,	89,	60,	64,	71,	77,	78,	80,	62,	61,	67,	69,	69,	57,	58,	69,	64,	61,	61,	69,	68,	66,	57,	64,	67,	64,	67,	60,	63,	64,	58,	67,	63,	50,	60,	43,	73,	58,	56,	59,	70,	57,	48,	76,	56,	58,	75,	64,	59,	54,	64,	68,	67,	56,	41,	59,	63,	75,	69,	47,	53,	62,	48,	71,	59,	47,	39,	47,	67,	51,	44,	40)

#make a dateframe
dat <- data.frame(route = route,
                  spp.rich = spp.rich)

```



## Distribution of species richness values

The distribution of species richness values looks like this
```{r, message=FALSE, warning=FALSE}

library(ggplot2)
qplot(spp.rich, 
      dat = dat)

```

A nice package for making graphs with ggplot is ggpubr.  This lots of "dependencies" so you'll see lots of red text fly by and it will take a minute to download.



```{r}
#download the packages
# install.packages("ggpubr")
# install.packages("dplyr")

library(ggpubr)
```

See http://www.sthda.com/english/rpkgs/ggpubr/ for more details on ggpubr.

A basic ggpubr histogram using the function gghistogram(). (Ignore the red text about "bind = 30")
```{r}
gghistogram(data = dat,
            x = "spp.rich")
```


Add the mean as a verticle line using the add = "mean" arguement
```{r}
gghistogram(data = dat,
            x = "spp.rich",
            add = "mean")
```


Change the "binwidth" 
```{r}
gghistogram(data = dat,
            x = "spp.rich",
            add = "mean",
            binwidth = 5)
```


Add a "rug" on the bottom of the plot where each datapoint is exactly.
```{r}
gghistogram(data = dat,
            x = "spp.rich",
            add = "mean",
            binwidth = 5,
            rug = TRUE)
```


Add a color to fill the bars.
```{r}
gghistogram(data = dat,
            x = "spp.rich",
            add = "mean",
            binwidth = 5,
            fill = "orange",
            rug = TRUE)
```




## Summary stats

* All your basic stats.
* Note meaninful for "route" b/c it is actually a categorical variable
```{r}
summary(dat)
```

## Accessing a column in a dataframe

Look at just the spp.rich column using $spp.rich
```{r}
dat$spp.rich
```



## Basic math stuff in R


### sum()

The summed of the spp richness of all routes.  Not really meaninful on its own.
```{r}
sum(dat$spp.rich)
```


## The min, max, etc.
```{r}

min(dat$spp.rich)

max(dat$spp.rich)

```


### Mean()

The mean number of species observed per route.
```{r}
mean(dat$spp.rich)
```


## The varinace var()

* On its own this value is hard to interpret.  But he larger the variance, the more variation there is between each route. 
* If all the routes had the exact same number of species, var = 0.0

```{r}
var(dat$spp.rich)
```


## The standard deviation (sd)

The standard deviation is a very common way to represent variability

```{r}
sd(dat$spp.rich)
```


## Variance vs. standard deviation

* The standard deviation is just the square root of the of the variance.
* We get the square root using sqrt()

```{r}
#SD using R's function
sd(dat$spp.rich)

#SD as the sqrt of the variance
sqrt(var(dat$spp.rich))

```


## Doing math for stats "by hand""


### Calculating the mean by hand

*The numerator*
```{r}
my.sum <- sum(dat$spp.rich)
```

*The denomintor*

Using the length() function to get the number of rows in the dataframe
```{r}
my.N <- length(dat$spp.rich)
```

**The mean**

Doing division using "/"

```{r}

my.mean <- my.sum/my.N
my.mean

```

**In one step**
Same thing to get the mean, just in 1 step
```{r}

my.mean <- sum(dat$spp.rich)/length(dat$spp.rich)
my.mean

```


**Checking our answer using == **
We can check our answer like this using the "==" function which asks R "are these two objects EXACTLy the same?"

```{r}
my.mean == mean(dat$spp.rich)
```


We can do the same thing to check that the standard deviation is indeed the square root of the mean

```{r}
sd(dat$spp.rich) == sqrt(var(dat$spp.rich))
```

If the were not the same, R would say "FALSE"


* Because R is very very very very precise any rounding errors will result in R saying that 2  things are NOT exactly the same.  Sometimes some R functions do  some rounding so you have to check "FALSE" answers sometimes.


## Calcualting variance & standard deviation by hand, step by step

* Variance and standard deviation are fundamental quantities in statistics.
* We'll step through each part of their calculation

### "Deviations" between the mean and each observation

Calculate the difference between each observation and the mean of all observation that you calculated above.

```{r}

Yi.deviations <- dat$spp.rich-my.mean

```



### Start making a dataframe (df)

* We can keep track of the math by making a spreadsheet-like object in R called a  dataframe using the data.frame command.  

```{r}
my.df <- data.frame(spp.rich = dat$spp.rich,
               Yi.deviations)

```


Look at the matrix.  Note that some of the deviations are positive and some are negative.

```{r}
head(my.df)
tail(my.df)
```


Look at a histogram of these values.  We won't focus on this now, but these values are called the "residuals"
```{r}
gghistogram(data = my.df,
            x = "Yi.deviations")
```




## Calculate the"Squared deviations" between the mean and each observation

Take your set of deviations and square them using "^2"

Here, we've done math on a whole list of numbers at the same time.

```{r}

Yi.deviations.square <- Yi.deviations^2

```


* Squaring is key b/c it makes deviations greater than the mean and less than the mean to have the same magnitude  
* That is, we go from "deviations" that are positive and negative to "squared deviations" that are all positive  


## Add the square deviations to the dataframe


* We can add stuff to our dataframe using the function cbind()
* cbind() stands for "column bind"

```{r}

my.df <- cbind(my.df, Yi.deviations.square)

```

Look at your expanding matrix

```{r}
head(my.df)
```


## Sum of squares between the mean & each deviation

* we can now calcualte the "sum of square deviations" 
* AKA the "sum of squares" (SS)
* This is the numerator (the thing on top) in the variance & standard deviation equations.*
* This is an important intermediary step in the process

```{r}

my.sum.of.squares<- sum(Yi.deviations.square)


```


This is a rather big number and I'm glad we don't have to calcualte it by hand : )

```{r}

my.sum.of.squares

```


## And now...the variance (var)

* We can now use the sum of squares(SS) to calcuate the variance. 
* This is is the SS divided by the sample size minus one. 
* That is SS/(n-1)
* Recall that we calcualted the sample size (N) above and put it in an object called "my.N".  
* We subtract 1 from N to get what is known as the "Degrees of freedom" (DF).  More on this later...

**Equation for the variance**
\[\sigma^2 = {\frac{\sum\limits_{i=1}^{n} \left(x_{i} - \bar{x}\right)^{2}} {n-1}}\]



```{r}
#The sample size
my.N

#degrees of freedom
dfs <- my.N - 1

#The var
my.var <- my.sum.of.squares/dfs

#Could also do it more directly
my.var <- my.sum.of.squares/(my.N - 1)


```


Check our var vs. R's var

```{r}
my.var == var(dat$spp.rich)
```

Got it!


## And now...the standard deviation (SD)

* The standard deviation is just the square root of the variance.
* We get the square root using the sqrt() function


**Equation for Standard Deviation**
Where x = a column of data in a dataframe
\[\sigma = \sqrt{\frac{\sum\limits_{i=1}^{n} \left(x_{i} - \bar{x}\right)^{2}} {n-1}}\]


**R-ish notation for SE**
\[SD = \sqrt{var(x)}\]


```{r}

my.sd <- sqrt(my.var)

```


We can check if our results are the same as R

```{r}
#Variance
var(dat$spp.rich)
my.var

#stdev
sd(dat$spp.rich)
my.sd

```


We can check this also like this usig "=="

```{r}

#stdev
sd(dat$spp.rich) == my.sd
```

## The standard error (SE)


* The standard error (SE) is a fundamental quantity in stats.  
* It is used as a measure of the precision of the data
* It is closely related to the concept of a 95% confidence interval (CI)
* Is is calculated the standard deviation divided by the square root of the sample size.

**Equation for Standard error (SE)**
Where x = a column of data in a dataframe
\[SE = \frac{SD}{\sqrt{N}}\]

**R-ish notation for SE**
\[SE = \frac{SD(x)}{sqrt(length(x))}\]


```{r}

my.se <- my.sd/sqrt(my.N)

```




## Calculating the SE directly

* For some reason the basic R set up doesn't have a function for the SE.
* There is a function for the SE, std.error, in the package plotrix

```{r}
#Download plotrix
# install.packages("plotrix")
library(plotrix)

std.error(dat$spp.rich)

```


## Advanced Pro trick: write your own function

* If there wasn't a function in for the SE, you could write your own like this.
* This is an advanced topic and is only shown for those who are curious.

```{r}
#with all the math
my.se.fnxn <- function(x){sqrt(var(x))/sqrt(length(x))}

#with less  math
my.se.fnxn <- function(x){sd(x)/sqrt(length(x))}


my.se.fnxn(dat$spp.rich)
```


## 95% Confidence intervals (95% CIs)

* The standard error is related closely to the 95% Confidence interval
* There are precise equations for calculating 95% CIs for different circumstances
* Roughtly, a 95% CIs extend on either side of the mean to indicate a plausible range of values the should contain the real value you are trying to estimate with your eamn
* As an approximatation, 95% CI is mean +/- 1.98*SE
* The upper bound of the CI is mean +  1.98*SE
* The lower bound of the CI is mean -  1.98*SE

```{r}
my.CI.upper <- my.mean + my.se*1.96
```






## Background info

https://www.r-bloggers.com/standard-deviation-vs-standard-error/




