---
title: 'Introduction to the to t-test: t-test'
author: "Nathan Brouwer | brouwern@gmail.com | @lobrowR"
date: "September 29017"
output:
  pdf_document: default
  html_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 4, 
                      fig.height = 2,
                      warning = FALSE,
                      message = FALSE)
#set rounding
options(digits = 4)

#set updata
brook.trout.PRES.ABS <- c(rep("present",6,),rep("absent",6))

salmon.released <- c(820,
                     960,700,545,769,1001,467,959,1029,27,998,936)
salmon.surv <- c(166,136,153,103,173,188,180,178,326,7,120,135)
salmon <- data.frame(brook.trout.PRES.ABS,
                     salmon.released,
                     salmon.surv)
percent.surv <-salmon$salmon.surv/salmon$salmon.released
salmon$percent.surv <- percent.surv

library(ggplot2)
library(ggpubr)
library(dplyr)

```


## Conducting a 2-sample t-test

* We'll use the main salmon dataframe we plotted previously.  
* All t-tests in R are done using the t.test function
* There are unfortunately a few different ways to can get t.test to run things for you; you might see some variation dependong on what resources you are looking at
* We'll do what is probalby the most common way, using the tilda formula notation
* This is: "response.variable ~ predictor.variable"
* Note we have to include "data = salmon" so R knows where to find the reponse and predictor variables

```{r}
t.test(percent.surv ~ brook.trout.PRES.ABS, 
       data = salmon)
```


## Saving output of t-test

We can save the output of a t test into an R object

```{r, echo = F}
t.out <- t.test(percent.surv ~ brook.trout.PRES.ABS, 
       data = salmon)
```


```{r, include=FALSE}
mean1 <- t.out$estimate[1]
mean2 <- t.out$estimate[2]
```

## Examining Output of a t-test

* Examine the output of the t-test.  Idetinfy the following thigns
* The t-statistics
* degrees of freedom (df)
* p-value


Note that the output reports an interesting thing, a **95% confidence interval.**

Specifically, this is the **95% for the difference between the two mean values.**  The means are `r mean1` and  `r mean2`, and their difference is `r mean1 - mean2`.  

The **null hypothesis (Ho)** is that the **difference between the 2 means**  is essentially equal to zero.  `r mean1 - mean2` is close to zero, and we use the t-test and confidence intervals to determine if a difference of `r mean1 - mean2` could just result from random sampling error.




