---
title: "Practicing T-test in R"
subtitle: "Part 2: 2-sample and paired t-tests"
author: "Nathan Brouwer|brouwern@gmail.com|@lobrowR"
editor_options:
  chunk_output_type: console
output:
  pdf_document: default
  html_document: default
date: "October 2017"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 4, 
                      fig.height = 2,
                      warning = FALSE,
                      message = FALSE)

library(ggplot2) #plotting tools
library(ggpubr)
library(dplyr)  #data cleaning tools
bodytemp <- read.csv("bodytemp.csv")

bodytemp$gender.MF <- recode(bodytemp$gender, 
                    "1" = "male", 
                    "2" = "female")
```

## Outline

* Part 2: 2-sample t-test & paired t-test
  + Question 2: Is the mean human body statistically different from 98.6?
  + Question 3: How do Male & Female Body Temps compare?
  + Question 4: How is a paired t-test similar to a 1-sample t-test?
  + OPTIONAL: Question 5: What is the relationship between resting heart rate & body temperature

## Question 3: How do Male and Female Body Temps compare?


### Graph 3: How do males and female compare?

An excellent way to compare data that is in two groups (eg male and female) is with a boxplot

```{r}
#library(ggpubr)
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF")
```


We can code by the color of the lines like this using color = 
```{r, eval=FALSE}
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF",
          color = "gender.MF")
```

And we can change the fill using fill (note, only do color or fill, not both)
```{r, , eval=FALSE}
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF",
          fill = "gender.MF")
```


We can improve the labels using xlab and ylab
```{r, eval=FALSE}
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF",
          fill = "gender.MF",
          xlab = "Gender",
          ylab = "Body temperature (F)")
```



Boxplots tell us a lot about the data, but they can be further improved by overlaying the raw data.  We can do this in ggpubr using add = "jitter"
```{r, eval=FALSE}
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF",
          fill = "gender.MF",
          xlab = "Gender",
          ylab = "Body temperature (F)",
          add = "jitter")
```




#### OPTIONAL: OLD SCHOOL - boxplot
```{r, eval=FALSE}
      boxplot(temp ~ gender.MF, data = bodytemp)
```


### Evaluting the graphs
Males appear to have a lower median temp than female.  Is this biologically based or could this difference just be due to chance (eg, just b/c we have a small sample)?


### 2-sample t-test

* We can assess this using a 2-sample t-test.  
* Before, we had all the data pooled together, hence we did a "1-sample" test. 
* Now we are splitting the data into two groups and therefore its a **two sample** test
* Two sample tests probably the most common type of t-test
* Therefore, if someones says that they did a t-test, the probably mean a 2-sample test
* For a 2-sample test we don't specify "mu"
* We specify the test using the equation notation just like the boxplot, using the tilde symbol (~)

```{r}
t.test(temp ~ gender.MF, 
       data = bodytemp)
```

* What does this test indicate about male and female body temp?
* Female body temp is higher - is it closer to 98.6?


## Plotting the 2-sample t-test

### Plot boxplots w/ p value

ggpubr has a cool feature for adding p-values to plots called stat_compare_means().  Remember the "+" between ggboxplot() and stat_compare_means()!

```{r}
ggboxplot(data = bodytemp,
          y = "temp",
          x = "gender.MF",
          fill = "gender.MF",
          xlab = "Gender",
          ylab = "Body temperature (F)") +
  stat_compare_means(method = "t.test")
```



### Plot means and error bars

We can quickly plot means and error bars using ggerrorplot().  The default is to plot means with error bars based on the standard error; we'll change this to 95% CIs by setting desc_stat = mean_ci

```{r}
ggerrorplot(data = bodytemp,
           desc_stat =  "mean_ci",
          y = "temp",
          x = "gender.MF")
```

Note that the error bars overlap a fair bit, even those the t-test gives us a value of p < 0.05.



### Plot means, error bars and raw data

A create type of plot is a "beeswarm plot", which ggpubr just calls a "dotplot".  We can use  add = "mean_ci" to add the means and CIs.  Note how tiny the CIs are relative to the full range of the data.

```{r}
ggdotplot(data = bodytemp,
          y = "temp",
          x = "gender.MF", 
          add = "mean_ci")
```


For comparison, change it to add = mean_sd.  What do you notice?

```{r, eval=FALSE}
ggdotplot(data = bodytemp,
          y = "temp",
          x = "gender.MF", 
          add = "mean_sd")
```


The SD is a measure of the variation in the data, while the SE/95% CI is a measure of how precisely we have measured the mean.  Here, the SD bars are wide b/c there is a lot of variation in the data, but the 95% CI are narrow b/c the sample size is pretty big. 


## Question 5: How is a paired t-test similar to a 1-sample t-test?

* A paired t-test is a very common type of t-test
* A common type of pairing is before and after a treatment occurred
* A paired treatment can be set up a couple different ways in R
* Mathematically, a paired t-test is similar to a 1-sample t-test.
* With a paired t-test we are not interested in whether the overall means of each group or time point (ie before, after) are different
* We are interested in whether the difference between each pair of measurements is consistently different from zero
* **A paired t-test is therefore equivalent to a 1-sample t-test where the population parameter mu=0**

### Simulate data

* We will look at paired t-tests with some fake data.  
* I've simulated some data representing a person's body temp before the experimental treatment occurred (before)
* I've also simulated body temps after a stress treatment occurs (stressed)
* The hypothesis (Ha) is that being stressed changes your body temp.
* The null hypothesis (Ho) is that there is no consistent impact of stress on body temp.
* That is, while some peoples body temp goes up after the treatment, other people's go down, and on average the change in temp is about 0.
* (NOTE: I intentionally made the means to be different, so the difference is not surprising!)

### Simulate fake paired data

This code makes the fake data; ignore it

```{r}
      # ignore this
      # lm.out <- lm(temp ~ heartrate, data = bodytemp)
      # temp.orig <- simulate(lm.out, seed = 100)
      # temp.stress <- simulate(lm.out, seed = 101) + 0.5+ rnorm(length(bodytemp), mean = 0, sd = sd(bodytemp$temp)/2)
      # new.bodytemp <- data.frame(before = temp.orig$sim_1,stressed = temp.stress$sim_1)
      # write.csv(new.bodytemp, "fake_paired_temp_data.csv",row.names = F)

```


### Load the fake paired temp data

Follow the steps use previously to load the data.  I'm going to use the command read.csv(), but you can do it however you want.

```{r}
new.bodytemp <- read.csv(file = "fake_paired_temp_data.csv")
```

### Calculate the difference between "before"

* We can do some simple math to calculate the difference between before and after
* we use the dollar sign ($) to select the columns we want

```{r}
difference1 <- new.bodytemp$before -  new.bodytemp$stressed
```


We can add this to our dataframe like this

```{r}
new.bodytemp$difference1 <- difference1
```


Note that we could do this all in 1 step if we wanted.  Lets flip the order of the values.

```{r}
new.bodytemp$difference2<- new.bodytemp$stressed - new.bodytemp$before 
```


### Visualize difference

Look at distribution of differences

```{r}
gghistogram(new.bodytemp, 
            x = "difference1",
            add = "mean")
```

What is the mean represent?


#### OPTIONAL - OLDSCHOOL
```{r, eval=FALSE}
    hist(new.bodytemp$difference1)
```


### 1-sample t-test on difference

* We'll complete the paired t-test process by conducting a 1-sample test, setting mu = 0.
* We are therefore testing the hypothesis that the average **difference** between before and stressed is greater than zero

```{r}
t.test(new.bodytemp$difference1 , 
       mu = 0)
```


### What's the difference between these two results?

* Our 1st difference column  difference1 was calculated as "new.bodytemp$before -  new.bodytemp$stressed"
* This is, "before" - "stressed"
* Our 2nd difference column  difference2 was calculated as " new.bodytemp$stressed - new.bodytemp$before"
* That is, "stressed - before"


Both of these columns contain equivalent data, just the signs (pos or neg) are different depending on what gets subtracted from what
```{r}
#raw data
mean(new.bodytemp$difference1)
mean(new.bodytemp$difference2)
```


We can over the top check this using the abs() command to get the absolute values
```{r}
#absolute values with abs()
mean(abs(new.bodytemp$difference1))
mean(abs(new.bodytemp$difference2))
```


Compare the 2 t-tests
```{r, eval=FALSE}
#difference1 column
t.test(new.bodytemp$difference1 , 
       mu = 0)

#difference2 column
t.test(new.bodytemp$difference2 , 
       mu = 0)
```

The results are the same in terms of the p-value but different in other ways.  Can you spot the differences?  There are 2 (2 types of differences, 3 numbers that are different)


#### Paired t-test the normal way

* Its actually not necessary to calculate the difference - the t.test function can do it on the fly.
* To do things directly, give t.test() two things: the before column and the stressed column
* then, tell t.test() "paired = TRUE"
* This tells t.test() that the two columns of data are directly paired; that is, each row of data contains two numbers that are paired.

```{r, eval=FALSE}
t.test(new.bodytemp$before,
       new.bodytemp$stressed,
       paired = TRUE)
```





## OPTIONAL Question 5:  Introducing scatter plots by considering the relationship between resting heart rate & temp

#### Making a scatterplot
```{r, eval=FALSE}
ggscatter(y = "temp",x = "heartrate",
          data = bodytemp)
```

There might be some relationship here.  We could explore this further if we wanted with linear regression.


### OPTIONAL-OLDSCHOOL: Making a scatterplot using plot()
```{r, eval=FALSE}
      plot(temp ~ heartrate,data = bodytemp)
```



### Males vs. Female

Do male and females have different relationships between resting hear rate and body temp?

```{r, eval=FALSE}
ggscatter(y = "temp",x = "heartrate",
          data = bodytemp,
          color = "gender.MF",
          shape = "gender.MF",
          add = "reg.line")
```


The patterns appear to be similar.  We could explore this further with multiple linear regression.


#### OPTIONAL-OLDSCHOOL
Subset the data by males and females
```{r}
males <- subset(bodytemp,  gender.MF == "male")
females <- subset(bodytemp,  gender.MF == "female")
```


Plot males and females separately
```{r, eval=FALSE}
par(mfrow = c(1,1))
#set xlims
xlims <- c(min(bodytemp$heartrate),max(bodytemp$heartrate))
plot(temp ~ heartrate, data = males,xlim = xlims)
points(temp ~ heartrate, data = females, col = "green")
```



## Appendices

#### Appendix 1: Code to rebuild the data from scratch

This code allows you to re-build the data without loading any data.

```{r}
temp <- c(96.3,96.7,96.9,97,97.1,97.1,97.1,97.2,97.3,97.4,97.4,97.4,97.4,97.5,97.5,97.6,97.6,97.6,97.7,97.8,97.8,97.8,97.8,97.9,97.9,98,98,98,98,98,98,98.1,98.1,98.2,98.2,98.2,98.2,98.3,98.3,98.4,98.4,98.4,98.4,98.5,98.5,98.6,98.6,98.6,98.6,98.6,98.6,98.7,98.7,98.8,98.8,98.8,98.9,99,99,99,99.1,99.2,99.3,99.4,99.5,96.4,96.7,96.8,97.2,97.2,97.4,97.6,97.7,97.7,97.8,97.8,97.8,97.9,97.9,97.9,98,98,98,98,98,98.1,98.2,98.2,98.2,98.2,98.2,98.2,98.3,98.3,98.3,98.4,98.4,98.4,98.4,98.4,98.5,98.6,98.6,98.6,98.6,98.7,98.7,98.7,98.7,98.7,98.7,98.8,98.8,98.8,98.8,98.8,98.8,98.8,98.9,99,99,99.1,99.1,99.2,99.2,99.3,99.4,99.9,100,100.8)
gender <- c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2)
heartrate <- c(70,71,74,80,73,75,82,64,69,70,68,72,78,70,75,74,69,73,77,58,73,65,74,76,72,78,71,74,67,64,78,73,67,66,64,71,72,86,72,68,70,82,84,68,71,77,78,83,66,70,82,73,78,78,81,78,80,75,79,81,71,83,63,70,75,69,62,75,66,68,57,61,84,61,77,62,71,68,69,79,76,87,78,73,89,81,73,64,65,73,69,57,79,78,80,79,81,73,74,84,83,82,85,86,77,72,79,59,64,65,82,64,70,83,89,69,73,84,76,79,81,80,74,77,66,68,77,79,78,77)

bodytemp <- data.frame(temp = temp,gender = gender,heartrate = heartrate)

```

